---
title: "chapter05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 05 

Loading data

```{r libraries, include=FALSE, echo=FALSE}
library('rethinking')
data("WaffleDivorce")
```
```{r}
d = WaffleDivorce
head(d)
```


### Standarize predictors

```{r standarize, echo=TRUE}
d$MedianAgeMarriage.s = (d$MedianAgeMarriage - mean(d$MedianAgeMarriage) ) / sd(d$MedianAgeMarriage)
```

### Fit the model

```{r}
m5.1 = map(
  alist(
    Divorce ~ dnorm( mu, sigma),
    mu <- a + bA * MedianAgeMarriage.s, 
    a ~ dnorm( 10, 10),
    bA ~ dnorm( 0, 1),
    sigma ~ dunif( 0, 10)
  ), 
  data = d
)
```
Compute the percentile interval of mean.

1. Vector with range of *MedianAgeMarriage.s* (standarized)

2. Obtain the means of sampled values

3. Interval of means

```{r , echo=TRUE, include=TRUE, tidy=TRUE, warning=FALSE, results= "hide"}
MAM.seq = seq( from = -3, to = 3.5, length.out = 30)
mu = link(m5.1, data = data.frame(MedianAgeMarriage.s = MAM.seq))
mu.PI = apply( mu, 2, PI)
```
Plotting the resulting model
```{r, newplot = TRUE}
{plot( Divorce ~ MedianAgeMarriage.s, data = d, col = rangi2)
abline( a = coef(m5.1)['a'] , b = coef(m5.1)['bA']  )
shade( mu.PI, MAM.seq)}
```
```{r}
precis(m5.1)
```

## Fitting a multivariate model

We are going to use two predictors (marriage rate and marriage median age) to model the divorce rate per state

Here is the fitting code:

```{r}
d$Marriage.s  = (d$Marriage - mean(d$Marriage) ) / sd(d$Marriage)

m5.3 = map(
  alist(
    Divorce ~ dnorm( mu, sigma), 
    mu <- a + bR * Marriage.s + bA * MedianAgeMarriage.s,
    a ~ dnorm( 10, 10),
    bR ~ dnorm( 0, 1),
    bA ~ dnorm( 0, 1),
    sigma ~ dunif(0,10)
  ), data = d
)
precis( m5.3)
```


We observe that the posterior mean for marriage rate (bR) is now close to zero with probability on both sides. Let's see that in a plot:


```{r}
plot( precis( m5.3) )
```

Interpretation:

*Once we know median age at marriage for a state (bA), there is little or no additional predictive power in also knowing the rate of marriage in that state.*


## Plotting multivariate posteriors

### Predictor residual plot

We have so far 2 predictors:

1. marriage rate *Marriage.s*

2. median age at marriage *MedianAgeMarriage.s*

To compute the predictor residuals for both of them we will use the other predictor to model it.

```{r}

m5.4 = map(
  alist(
    Marriage.s ~ dnorm( mu, sigma ),
    mu <- a + b * MedianAgeMarriage.s , 
    a ~ dnorm( 0, 10 ),
    b ~ dnorm( 0, 1 ),
    sigma ~ dunif( 0, 10)
  ), 
  data = d
)
```

To compute the residuals:

1. Compute the expected value at MAP for each state

2. Compute residual for each state

```{r}
mu = coef( m5.4 )['a'] + coef( m5.4 )[ 'b' ] * d$MedianAgeMarriage.s
m.resid = d$Marriage.s - mu # actual value - predicted value
```

Plotting the relationship

```{r}
{ plot( Marriage.s ~ MedianAgeMarriage.s, d, col = rangi2)
  abline( a = coef( m5.4)['a'] , b = coef(m5.4)['b'] )
  # loop over states
  for (i in 1:length(m.resid)){
    x = d$MedianAgeMarriage.s[i] # location of line segment
    y = d$Marriage.s[i] # observed end point of line segment (actual value)
    lines( c(x,x) , c( mu[i], y) , lwd = 0.5 , col = col.alpha("black", 0.5) )
  }
}
```

