---
title: "chapter06"
author: "Pedro Sfriso"
date: "20/04/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 06 Exercices

#### Hard 

Preparing the models

```{r loading}
library(rethinking)
data("Howell1")
d = Howell1

# standarizing age 
d$age = (d$age - mean(d$age) )/sd(d$age)

set.seed( 1000 )
i = sample( 1:nrow(d), size = nrow(d)/2 )
d1 = d[  i, ]
d2 = d[ -i, ]
cat( dim(d1), dim(d2) )
```

We have randomly formed two data frames, each with 272 rows. This will be our test and training set.

#### Building the models

```{r models}

m1 = alist(
  height ~ dnorm(mean = mu, sd = sigma),
  mu <- a + b1 * age,
  c(a, b1) ~ dnorm(mean = 0, sd = 20),
  sigma ~ dunif(min = 0, max = 50)
)

m2 = alist(
  height ~ dnorm(mean = mu, sd = sigma),
  mu <- a + b1 * age + b2 * age^2,
  c(a, b1, b2) ~ dnorm(mean = 0, sd = 20),
  sigma ~ dunif(min = 0, max = 50)
)

m3 = alist(
  height ~ dnorm(mean = mu, sd = sigma),
  mu <- a + b1 * age + b2 * age^2 + b3 * age^3,
  c(a, b1, b2, b3) ~ dnorm(mean = 0, sd = 20),
  sigma ~ dunif(min = 0, max = 50)
)

m4 = alist(
  height ~ dnorm(mean = mu, sd = sigma),
  mu <- a + b1 * age + b2 * age^2 + b3 * age^3 + b4 * age^4,
  c(a, b1, b2, b3, b4) ~ dnorm(mean = 0, sd = 20),
  sigma ~ dunif(min = 0, max = 50)
)

m5 = alist(
  height ~ dnorm(mean = mu, sd = sigma),
  mu <- a + b1 * age + b2 * age^2 + b3 * age^3 + b4 * age^4 + b5 * age^5,
  c(a, b1, b2, b3, b4, b5) ~ dnorm(mean = 0, sd = 20),
  sigma ~ dunif(min = 0, max = 50)
)

m6 = alist(
  height ~ dnorm(mean = mu, sd = sigma),
  mu <- a + b1 * age + b2 * age^2 + b3 * age^3 + b4 * age^4 + b5 * age^5 + b6 * age^6,
  c(a, b1, b2, b3, b4, b5, b6) ~ dnorm(mean = 0, sd = 20),
  sigma ~ dunif(min = 0, max = 50)
)

```

#### Fitting the models

First set reasonable starting conditions:

```{r}
a.start <- mean(d$height)
sigma.start <- sd(d$height)
```

Model fitting

```{r model_fitting}

fit.m1 <- map( flist = m1, data = d1, start = list(a = a.start, sigma = sigma.start, b1 = 0))

fit.m2 <- map( flist = m2, data = d1, start = list(a = a.start, sigma = sigma.start, b1 = 0, b2 = 0))

fit.m3 <- map( flist = m3, data = d1, start = list(a = a.start, sigma = sigma.start, b1 = 0, b2 = 0, b3 = 0))

fit.m4 <- map( flist = m4, data = d1, start = list(a = a.start, sigma = sigma.start, b1 = 0, b2 = 0, b3 = 0, b4 = 0))

fit.m5 <- map( flist = m5, data = d1, start = list(a = a.start, sigma = sigma.start, b1 = 0, b2 = 0, b3 = 0, b4 = 0, b5 = 0))

fit.m6 <- map( flist = m6, data = d1, start = list(a = a.start, sigma = sigma.start, b1 = 0, b2 = 0, b3 = 0, b4 = 0, b5 = 0, b6 = 0))

```

###### 6H1: Model comparison.

```{r}
# WAIC is the default choice
compare(fit.m1, fit.m2, fit.m3, fit.m4, fit.m5, fit.m6)
```

###### 6H2: Model plotting

Common settings.

```{r}
n.trials <- 1e4
age.seq <- seq(from = -2, to = 3.5, length.out = 58)
prediction.data <- data.frame(age = age.seq)
```

And basic functions not shown.

```{r, echo=FALSE, include=FALSE}

computeMu <- function(model, data, n.trials) {
  mu <- link(fit = model, data = data, n = n.trials)
  return(mu)
}

computeMuMean <- function(mu) {
  mu.mean <- apply(X = mu, MARGIN = 2, FUN = mean)
  return(mu.mean)
}

computeMuPI <- function(mu) {
  mu.PI <- apply(X = mu, MARGIN = 2, FUN = PI, prob = .97)
  return(mu.PI)
}

simulateHeights <- function(model, prediction.data) {
  simulated.heights <- sim(fit = model, data = prediction.data)
  return(simulated.heights)
}

plotResults <- function(model, prediction.data, original.data, n.trials) {
  mu <- computeMu(model, prediction.data, n.trials)
  mu.mean <- computeMuMean(mu)
  mu.PI <- computeMuPI(mu)
  simulated.heights <- simulateHeights(model = model, prediction.data = prediction.data)
  simulated.heights.PI <- apply(X = simulated.heights, MARGIN = 2, FUN = PI)
  plot(height ~ age, data = original.data, col = rangi2)
  lines(x = prediction.data$age, y = mu.mean, lty = 2)
  lines(x = prediction.data$age, y = mu.PI[1,], lty = 2)
  lines(x = prediction.data$age, y = mu.PI[2,], lty = 2)
  shade(object = simulated.heights.PI, lim = prediction.data$age)
}
```

```{r}
plotResults(model = fit.m1, prediction.data = prediction.data, original.data = d1, n.trials = n.trials)
```
```{r}
plotResults(model = fit.m2, prediction.data = prediction.data, original.data = d1, n.trials = n.trials)
```
```{r}
plotResults(model = fit.m3, prediction.data = prediction.data, original.data = d1, n.trials = n.trials)
```



```{r}
plotResults(model = fit.m4, prediction.data = prediction.data, original.data = d1, n.trials = n.trials)
```


```{r}
plotResults(model = fit.m5, prediction.data = prediction.data, original.data = d1, n.trials = n.trials)
```



```{r}
plotResults(model = fit.m6, prediction.data = prediction.data, original.data = d1, n.trials = n.trials)
```

```{r}
precis(fit.m6)
```


```{r}
comparison = compare(fit.m1, fit.m2, fit.m3, fit.m4, fit.m5, fit.m6)
plot(comparison)
```

```{r, echo=FALSE}
weights = comparison@output$weight

# simulate heights

simulated.heights.m1 <- simulateHeights(model = fit.m1, prediction.data = prediction.data)

simulated.heights.m2 <- simulateHeights(model = fit.m2, prediction.data = prediction.data)

simulated.heights.m3 <- simulateHeights(model = fit.m3, prediction.data = prediction.data)

simulated.heights.m4 <- simulateHeights(model = fit.m4, prediction.data = prediction.data)

simulated.heights.m5 <- simulateHeights(model = fit.m5, prediction.data = prediction.data)

simulated.heights.m6 <- simulateHeights(model = fit.m6, prediction.data = prediction.data)

```
```{r}
sim.heights =   weights[6]* simulated.heights.m1 +
                weights[5]* simulated.heights.m2 +
                weights[4]* simulated.heights.m3 +
                weights[1]* simulated.heights.m4 +
                weights[2]* simulated.heights.m5 +
                weights[3]* simulated.heights.m6
```

Plot results

```{r}
simulated.heights.ensemble.PI <- apply( X = sim.heights, MARGIN = 2, FUN = PI)
```

```{r}
{plot(height ~ age, data = d1, col = rangi2)
shade(object = simulated.heights.ensemble.PI, lim = prediction.data$age )}
```


##### Compute deviance for the test sample

```{r}

## model 1
coefs = coef(fit.m1)
a = coefs["a"]
b1 = coefs["b1"]
sigma = coefs["sigma"]
mu <- a + b1*d2$age 
log.likelihood <- sum( dnorm(x = d2$height, mean = mu, sd = sigma, log = TRUE) )
dev.m1 <- -2*log.likelihood

## model 2
coefs = coef(fit.m2)
a = coefs["a"]
b1 = coefs["b1"]
b2 = coefs["b2"]
sigma = coefs["sigma"]
mu <- a + b1*d2$age + b2*(d2$age)^2
log.likelihood <- sum( dnorm(x = d2$height, mean = mu, sd = sigma, log = TRUE) )
dev.m2 <- -2*log.likelihood

## model 3
coefs = coef(fit.m3)
a = coefs["a"]
b1 = coefs["b1"]
b2 = coefs["b2"]
b3 = coefs["b3"]
sigma = coefs["sigma"]
mu <- a + b1*d2$age + b2*(d2$age)^2 + b3*(d2$age)^3 
log.likelihood <- sum( dnorm(x = d2$height, mean = mu, sd = sigma, log = TRUE) )
dev.m3 <- -2*log.likelihood


## model 4
coefs = coef(fit.m4)
a = coefs["a"]
b1 = coefs["b1"]
b2 = coefs["b2"]
b3 = coefs["b3"]
b4 = coefs["b4"]
sigma = coefs["sigma"]
mu <- a + b1*d2$age + b2*(d2$age)^2 + b3*(d2$age)^3 + b4*(d2$age)^4 
log.likelihood <- sum( dnorm(x = d2$height, mean = mu, sd = sigma, log = TRUE) )
dev.m4 <- -2*log.likelihood


## model 5
coefs = coef(fit.m5)
a = coefs["a"]
b1 = coefs["b1"]
b2 = coefs["b2"]
b3 = coefs["b3"]
b4 = coefs["b4"]
b5 = coefs["b5"]
sigma = coefs["sigma"]
mu <- a + b1*d2$age + b2*(d2$age)^2 + b3*(d2$age)^3 + b4*(d2$age)^4 + b5*(d2$age)^5 
log.likelihood <- sum( dnorm(x = d2$height, mean = mu, sd = sigma, log = TRUE) )
dev.m5 <- -2*log.likelihood

## model 6
coefs = coef(fit.m6)
a = coefs["a"]
b1 = coefs["b1"]
b2 = coefs["b2"]
b3 = coefs["b3"]
b4 = coefs["b4"]
b5 = coefs["b5"]
b6 = coefs["b6"]
sigma = coefs["sigma"]
mu <- a + b1*d2$age + b2*(d2$age)^2 + b3*(d2$age)^3 + b4*(d2$age)^4 + b5*(d2$age)^5 + b6*(d2$age)^6
log.likelihood <- sum( dnorm(x = d2$height, mean = mu, sd = sigma, log = TRUE) )
dev.m6 <- -2*log.likelihood

```

```{r}
# echo deviances
r1 = c('dev.m1:', dev.m1)
r2 = c('dev.m2:', dev.m2)
r3 = c('dev.m3:', dev.m3)
r4 = c('dev.m4:', dev.m4)
r5 = c('dev.m5:', dev.m5)
r6 = c('dev.m6:', dev.m6)
rbind(r1,r2,r3,r4,r5,r6)
```

### 6H6

Use the following model with more strongly regularizing priors on the coefficients.

Fit the model using d1.

```{r}
m6.reg = alist(
  height ~ dnorm(mean = mu, sd = sigma),
  mu <- a + b1 * age + b2 * age^2 + b3 * age^3 + b4 * age^4 + b5 * age^5 + b6 * age^6,
  c(a, b1, b2, b3, b4, b5, b6) ~ dnorm(mean = 0, sd = 5),
  sigma ~ dunif(min = 0, max = 20)
)

fit.m6.reg =  map(flist = m6.reg, data = d1, start = list(a = 156.42, sigma =  8.17, b1 = 0, b2 = 0, b3 = 0, b4 = 0, b5 = 0, b6 = 0))

```

Reporting MAP for the model parameters

```{r}
precis( fit.m6.reg)
```

```{r}
plot( coeftab(fit.m6,fit.m6.reg) , pars=c("b1","b2","b3","b4","b5","b6") )
```


```{r}
plotResults(model = fit.m6.reg, prediction.data = prediction.data, original.data = d1, n.trials = n.trials)
```

Out of sample deviance

```{r}
# compute out-of-sample deviance using data in d2
coefs = coef(fit.m6.reg)
a = coefs["a"]
b1 = coefs["b1"]
b2 = coefs["b2"]
b3 = coefs["b3"]
b4 = coefs["b4"]
b5 = coefs["b5"]
b6 <- coefs["b6"]
sigma <- coefs["sigma"]
mu <- a + b1 * d2$age + b2 * (d2$age)^2 + b3 * (d2$age)^3 + b4 * (d2$age)^4 + b5 * (d2$age)^5 + b6 * (d2$age)^6
log.likelihood <- sum( dnorm(x = d2$height, mean = mu, sd = sigma, log = TRUE) )
out.of.sample.deviance <- -2*log.likelihood

out.of.sample.deviance
```


